Modeling Interactions of Ions with Ether- and Ester-linked Phospholipids

### 2019-09-25 paper.tex
Finalish

\documentclass[12pt,preprint,times]{article}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{flafter}
%\usepackage{floatflt}
\usepackage{placeins}
\usepackage{siunitx}
\usepackage{graphicx}
\graphicspath{{./figures/}}% Include figure files
%\usepackage{epstopdf}
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{appendix}
\usepackage{amsmath}
\usepackage{cases}
\usepackage{calc}
\usepackage{amssymb}
\usepackage[dvips]{color}
\usepackage{color}
\usepackage{enumitem}
\usepackage{soul}
\usepackage{indentfirst}
\usepackage{hyphenat}
\usepackage{xspace}
\usepackage{subcaption}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{tabularx}
\usepackage{xcolor}
\usepackage{lineno}
\usepackage{setspace}
\newcommand{\ra}[1]{\renewcommand{\arraystretch}{#1}}
%\newcommand{\cmpersec}{$\frac{\text{cm}}{\text{s}}$~}
\newcommand{\cmpersec}{cm/s~}
\newcommand{\etal}{\textit{et al.}}
\newcommand{\na}{Na\textsuperscript{+}~}
\newcommand{\cl}{Cl\textsuperscript{-}~}
\newcommand{\PM}{$\pm$~}
\newcommand{\invnm}{$\text{\si{\nm}}^{-1}$~}
\newcommand{\nmeter}{\si{\nm}~}
\renewcommand{\arraystretch}{1.8}
\newcommand{\TODO}{\hl{TODO}~}
\newcommand{\about}{$\sim$}
\newcommand{\aangstroms}{{\aa}ngstr{\"o}ms}
\newcommand{\aangstrom}{{\aa}ngstr{\"o}m}
\newenvironment{hle}{\color{red}}

@@

\begin{document}
\bibliographystyle{plain}
\title{Need a title!!!! Interaction of salt with ether- and ester-linked phospholipid bilayers}
\author{Matthew Saunders} 
\maketitle
\tableofcontents
\listoffigures
\end{document}

>>>>

\documentclass[12pt,openany,final]{book}
\usepackage[letterpaper,margin=1in,footskip=0.25in]{geometry}
\fontfamily{cmr}
\usepackage[final]{changes}
\usepackage{lipsum}
\usepackage[nottoc]{tocbibind}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{flafter}
\usepackage{floatrow}
%\usepackage{floatflt}
\usepackage{placeins}
\usepackage{siunitx}
\usepackage{graphicx}
\graphicspath{{figures/}}% Include figure files
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{appendix}
\usepackage{amsmath}
\usepackage{cases}
\usepackage{calc}
\usepackage{amssymb}
\usepackage{color}
\usepackage{enumitem}
\usepackage{soul}
\usepackage{indentfirst}
\usepackage{hyphenat}
\usepackage{xspace}
\usepackage{subcaption}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{tabularx}
\usepackage{xcolor}
\usepackage{lineno}
\usepackage{setspace}
\usepackage{titlesec}
\usepackage{hyperref}
\usepackage[superscript]{cite}
\usepackage{lineno}
\titleformat
{\chapter} % command
[display] % shape
{\sc\LARGE} % format
{Chapter \thechapter} % label
{0.2ex} % sep
{
} % before-code
\titlespacing*{\chapter}{0pt}{1in}{0pt}
\titlespacing*{\section}{0pt}{0pt}{0pt}
\hbadness=99999
%\epstopdfDeclareGraphicsRule{.tga}{png}{.png}{
%    convert #1 \OutputFile
%}
\hypersetup{
    colorlinks,
    citecolor=black,
    filecolor=black,
    linkcolor=black,
    urlcolor=black
}
%\newcommand{\cmpersec}{$\frac{\text{cm}}{\text{s}}$~}
\newcommand{\cmpersec}{cm/s~}
\newcommand{\etal}{\textit{et al.}}
\newcommand{\na}{Na\textsuperscript{+}~}
\newcommand{\cl}{Cl\textsuperscript{-}~}
\newcommand{\PM}{$\pm$~}
\newcommand{\invnm}{$\text{\si{\nm}}^{-1}$~}
\newcommand{\nmeter}{\si{\nm}~}
\renewcommand{\arraystretch}{1.8}
%\newcommand{\hl}{\added}
%\newcommand{\sout}{\deleted}
\newcommand{\TODO}{\hl{TODO}~}
\newcommand{\about}{$\sim$}
\newcommand{\aangstroms}{{\aa}ngstr{\"o}ms}
\newcommand{\aangstrom}{{\aa}ngstr{\"o}m}
\newenvironment{hle}{\color{red}}

@@

\author{Matthew Saunders}
\title{Modeling Interactions of Ions with Ether- and Ester-linked Phospholipids}
\begin{document}
\pagestyle{plain}
%\maketitle
\begin{titlepage}
\begin{centering}
~\\[1in]
Modeling of Interaction of Ions \\[2\baselineskip] with Ether- and Ester-linked phospholipids
~\\[3\baselineskip]
by
~\\[3\baselineskip]
Matthew W. Saunders
~\\[4\baselineskip]
A thesis submitted in partial fufillment\\
of the requirements of the degree of\\
Master of Science in Biology \\ with a concentration in Cell \& Molecular Biology\\
Department of Cell Biology, Microbiology and Molecular Biology\\
College of Arts and Sciences\\
University of South Florida\\
~\\[2\baselineskip]
Major Professor: Sameer Varma, Ph.D.\\
Sagar Pandit, Ph.D.\\
H. Lee Woodcock, Ph.D.\\
Jianjun Pan, Ph.D.\\
~\\[2\baselineskip]
%Date of Approval:\\
%XXX\\
~\\[3\baselineskip]
\end{centering}
\end{titlepage}
\doublespacing
\tableofcontents
\pagenumbering{roman}
\listoftables
\listoffigures
\bibliographystyle{elsarticle-num}
%\linenumbers
\doublespacing
\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}

Polar lipids, or phospholipids, are present in all parts of cells
and are used in many signalling and structural roles.
As structural molecules they act as the main component of cellular plasma membranes --- bilayers
of polar lipids that separate cells from their environments. Bilayer properties are heavily
influenced by the structure of their component polar lipids, and different lipids
are found in different organisms. 
A distinguishing feature of Archaeal plasma membranes is that their phospholipids contain ether-links, as opposed to bacterial and eukaryotic plasma
membranes where phospholipids primarily contain ester-links. In our work we examine the effects of
salt on bilayer structure in the case of an ether-linked lipid bilayer. We use molecular dynamics simulations and compare equilibrium properties of two model lipid
bilayers in NaCl salt solution --- POPC and its ether-linked analog that we refer to as HOPC. We make the following key observations.
The headgroup region of HOPC ``adsorbs'' fewer ions compared to the headgroup region of POPC. Consistent with this,
we note that the Debye screening length in the HOPC system is \about 10\% shorter than that
in the POPC system. Herein, we introduce a protocol to identify the lipid-water interfacial boundary that
reproduces the bulk salt distribution consistent with
Gouy-Chapman theory. We also note that the HOPC bilayer has excess solvent
in the headgroup region when compared to POPC,
coinciding with a trough in the electrostatic potential. Waters in
this region have longer autocorrelation times and smaller lateral diffusion rates compared to the corresponding region
in the POPC bilayer, suggesting that the waters in HOPC are more strongly
coordinated to the lipid headgroups. Furthermore, we note that it is this
region of tightly coordinated waters in the HOPC system that has a lower density of Na$^+$ ions.
Based on these observations we conclude that an ether-linked lipid bilayer has a lower binding
affinity for Na$^+$ compared to an ester-linked lipid bilayer.

\chapter{Introduction}
\linenumbers
\pagenumbering{arabic}

\section{Ether- and Ester-linked Phospholipids}
Phospholipids are a diverse species of biological molecules used in many roles in life~\cite{van:2008:lipidvariety}. These molecules
are amphipathic, with a headgroup that is hydrophillic attached to hydrophobic acyl chains by a
glycerol backbone~\cite{pandit:2008:simulationtextbook,israelachvili:2011:intermol}. 
These molecules, when above a critical concentration in aqueous solution, self assemble into
various aggregate structures in order 
to reduce the exposure of the hydrophobic chains
to polar solvent --- these structures include
micelles, vesicles, and bilayers along with others~\cite{israelachvili:2011:intermol}. 
Phospholipids tend to form bilayers, 
where the hydrophillic headgroups face the solvent and the 
hydrophobic acyl chains face into the center 
of the structure~\cite{israelachvili:2011:intermol,ashrafuzzaman:2012:membrane}. 
An image of a bilayer of a commonly studied
model lipid can be seen
in figure~\ref{fig:boximage}.
\begin{figure}[p]
    \caption[Image of a model bilayer of phospholipids from a simulation.]{ Image of a model bilayer of phospholipids from a simulation. The lipids can be seen
in the center of the image, with the hydrophobic chains colored in
light purple. The polar headgroup can be seen highlighted by the phosphate group in orange and the choline nitrogen in blue
facing the solvent. Solvent molecules have been removed for clairity. Figure generated
using VMD~\cite{vmd}}
\label{fig:boximage}
\includegraphics[width=\textwidth]{images/popc_image}
\end{figure}
In living cells, lipid bilayers act as the delimiter between the cell and the environment, acting as one of the major constituients of the cellular 
plasma membrane~\cite{reece:2014:biobook}. The plasma membrane consists of a mixture of many lipid species, as
well as proteins and lipid-like molecules, and the mixture is tuned by the cell to adapt the 
structure and flexibility of the plasma membrane to different environments~\cite{pandit:2008:simulationtextbook,van:2008:lipidvariety,zhang:2008:memhomeo}. 
Furthermore, the cellular membrane is the interface for exchange
of nutrients with the environment, and also the interface for the transmission of signals to and from the outside of the cell. This
exchange is done through passive diffusion, proteins and other molecules embedded into the membrane, and through the formation
of vesicles and buds out of the lipid bilayer~\cite{reece:2014:biobook}. The behavior and function of the plasma membrane
is heavily dependent on the structure of the bilayer, which again can be influenced directly by the mixture of phospholipids constituting the
membrane~\cite{van:2008:lipidvariety,zhang:2008:memhomeo}.

Phospholipids can be modified in a variety of ways to change their chemical construction,
and the structure of the individual molecules manifest significant changes
in the stucture of aggregates, like lipid bilayers or cell membranes.
Polar headgroups can be charged or zwitterionic~\cite{israelachvili:2011:intermol}. The fatty acid chains can be saturated or unsaturated, 
and can have many lengths depending 
on the needs of the cell --- differing the acyl chains influences the 
phase behavior of the bilayer, which can help cells to adapt to changes in
temperature~\cite{ashrafuzzaman:2012:membrane,zhang:2008:memhomeo}.

Phospholipids can also use several types of chemical links between the glycerol
backbone and the fatty acid chains~\cite{sehgal:1962, koga:2014}.
Phospholipids containing ether-links represent one of the many major evolutionary divides -- 
while they make up much of the phospholipids of archaeal membranes,
 they are present in only small fractions in eukarya and bacteria~\cite{sehgal:1962, koga:2014, kates:1993:archealipids}. 
These phospholipids differ from the more common ester-linked phospholipids of bacterial and 
eukaryotic membranes in that they lack fatty acid carbonyl groups (see figure \ref{fig:struc}).

\begin{figure}[p]
\caption[Chemical structures of representative phospholipids highlighting the difference
between ether- and ester-links.]{ 
Chemical structures of representative phospholipids highlighting the difference 
between ether- and ester-links. POPC (1-palmitoyl-2-oleoyl-sn-glycero-3-phosphatidylcholine) is a 
diester lipid and HOPC (1-hexadecyl-2-(9-octadecenyl)-sn-glycero-3-phosphatidylcholine) is its diether analog that lacks fatty acid carbonyl groups.
}
\label{fig:struc}
\includegraphics[height=0.66\textheight]{images/Structure_annotated}
\end{figure}

The chemical difference between ester- and ether-linked lipids has been shown to influence 
bilayer properties including both structure and permeability~\cite{guler:2009,jansen:1995, gawrisch:1992,haas:1990,fogarty:2015,kruczek:2017:ether}. 
Experiments show that model bilayers composed exclusively of ether-linked lipids have smaller dipole potentials compared to 
bilayers composed of their ester-linked analogs~\cite{gawrisch:1992}; a result also predicted by a molecular dynamics (MD) 
study done by our group~\cite{kruczek:2017:ether}. The interfacial water in ether-linked lipid bilayers is also more structured compared to that 
in ester-linked lipid bilayers, as observed in NMR experiments in the form a larger quadrupolar splitting constants~\cite{gawrisch:1992}, and as 
also reproduced in our simulations~\cite{kruczek:2017:ether}. Two independent experimental studies also show 
that ether-linked lipid bilayers are less permeable to water compared to bilayers composed of their ester-linked 
analogs~\cite{jansen:1995, guler:2009}.

Nevertheless, all studies aimed at examining differences between ester- and ether-linked lipid bilayers have been carried out in pure water, except for the one recent study by 
Leonard \etal~\cite{leonard:2018}. As such, several studies show that salt significantly 
affects the properties of ester-linked lipid bilayers~\cite{kruczek:2019,kruczek:2017, duro:2016,pabst:2007,sachs:2004,petrache:2006:swelling}. 
For example, in our previous works on the interactions of monovalent and divalent ions on zwitterionic ester-linked lipid bilayers~\cite{kruczek:2019,kruczek:2017}
we found that monovalent cations insert into headgroup region. 
We reported reduction in area per lipid, increase in  bilayer thickness, increased range of ordered solvent at the bilayer surface, and an increase in
the electrostatic dipole potential of the bilayer.
Despite the work done on these lipids in the past, the question of how ions interact with bilayers of ether-linked
phospholipids compared to ester-linked remains a topic of interest.

%\section{Models of Ion-Membrane Interactions - Poission-Boltzmann Theory}
%Ion-bilayer interactions are often modelled in the context of Poisson-Boltzmann theory, which
%expresses ions as a uncorrelated gas, moving through a dielectric continuum~\cite{israelachvili:2011:intermol}.

\section{Using Molecular Dynamics to Study Biological Membranes}
In order to explore the question of how ions interact with bilayers of ether- and ester-linked lipids, we utilize 
Molecular Dynamics (MD) simulations.

Computational methods like molecular often have to strike a balance between accuracy and computational feasibility --- approximations
to the interactions in a system can significantly reduce computational cost, and if carefully done can still result in
accurate reproduction of experimental results~\cite{pandit:2008:simulationtextbook}.

Molecular dynamics approximates systems by treating particles classically. Particles are expressed as points in space, 
with masses centered at the nucleus~\cite{pandit:2008:simulationtextbook}.
Functional forms and terms to calculate the potential energy of systems are defined in a force-field.
Bonds and bond angles in molecules are modeled as harmonic potentials, and torisional parameters within molecules are modeled
using periodic functions. Non-bonded interactions
are modeled using simple pairwise-additive potentials and in general do not treat many-body interactions~\cite{tieleman:1997:memsim,pandit:2008:simulationtextbook}. 
Van der Waals (VdW) dispersion and hard-shell repulsion are usually treated using a Lennard-Jones 6-12 potential,
and short-range electrostatics are represented by a coulomb potential~\cite{tieleman:1997:memsim,pandit:2008:simulationtextbook}.
The potential terms are added together over the whole system of particles
to get the overall potential function of the system, which is then used to evolve the system over time by integrating
Newton's laws of motion~\cite{frenkel:2001:molsimbook}. An example of the total potential can be seen in equation~\ref{eq:mdpot}~\cite{pandit:2008:simulationtextbook}.
\begin{equation}
    \label{eq:mdpot}
\begin{aligned}
    V_{tot}=&\sum_{\text{bonds}}K_b(r-r_0)^2 \\+& \sum_{\text{angles}}K_{\theta}(\theta - \theta_0)^2 \\+& \sum_{\text{improper}}K_\Phi(\Phi-\Phi_0)^2
    \\+& \sum_{\text{torision}}K_\phi[1-cos(n\phi-\phi_0)] \\+& \sum_{\text{Estatic}}\frac{q_iq_j}{r_{ij}} \\+& \sum_{VdW}\frac{A_{ij}}{r^{12}_{ij}}-\frac{B_{ij}}{r^6_{ij}}
\end{aligned}
\end{equation}

Most force-fields for molecular dynamics treat charges on particles in the system as an average picture of overall electronic behavior, as the
timescale of electronic fluctuations is much smaller than the timescale of MD simulations; this means bonds are permanent,
and average charge densities are fixed to particles~\cite{pandit:2008:simulationtextbook}.
Some force fields treat electronic polarizability as an explicit term in the potential function~\cite{li:2017:drude,ponder:2010:current}; however in general
MD force-fields do not include this in order to improve computation speed, and it has been found that
~\cite{pandit:2008:simulationtextbook}.

To ensure accurate reproduction of experimental results from complex systems, 
force-fields are developed and optimized to directly reproduce experimental data
in simple model systems. To develop force-fields, systems of small molecules can be 
optimized to reproduce experimental heats and enthalpies of 
vaporization, molecular volumes, 
and densities of condensed phases~\cite{berger:1997,chiu:1999:optimization,chiu:2003:structure,chiu:2009}.
For systems of whole lipid bilayers, experimental methods such as deuterium NMR, small angle x-ray scattering, and small angle neutron scattering 
can be used to get ensemble average values for bilayer ordering and structure and are often used to 
validate lipid simulations~\cite{pandit:2008:simulationtextbook,fogarty:2015,chiu:2009,kruczek:2017:ether,kruczek:2017,leonard:2018,li:2017:drude,nagle:2000}.
We can also look at electrostatic properties such as the bilayer electrostatic dipole potentials~\cite{gawrisch:1992,berkowitz:2006:aqueous}, or mechanical
properties like bilayer compressibility or surface tension~\cite{kruczek:2017:ether,kruczek:2017,jahnig:1996:surface}.
For optimization of force-fields using small molecules, experimental heats and enthalpies of vaporization, molecular volumes, 
and densities of condensed phases can be usedv~\cite{berger:1997,chiu:1999:optimization,chiu:2003:structure,chiu:2009}.

The earliest lipid simulations often used a united atom representation of the lipids, which simplified lipid chains by expressing carbon groups
as beads rather than explicitly defining hydrogens~\cite{raghavan:1992,marrink:1993,egberts:1994}. These simulations focused on validation with experimental
values for bilayer structure, such as area per lipid and bilayer thickness that are taken from SAXS and SANS measurements. 
Later development of force-fields included calculating partial charges
for different parts of the lipid molecules using \emph{ab initio} quantum calculations~\cite{chiu:1995:incorporation,marrink:1993}. Chiu \etal
calculated charges in this way for the GROMOS force-field of DPPC in 1995~\cite{chiu:1995:incorporation}.

Berger \etal sought to further improve the densities given by the GROMOS force-field in 1997 
by optimizing VdW parameters using molecular volumes and
heats of vaporization in small hydrocarbon molecules to improve the acyl chains of DPPC~\cite{berger:1997},
work that was continued by Chiu \etal~\cite{chiu:1999:optimization,chiu:2003:structure}, 
Later, further optimization was carried out by Chiu \etal 
in 2009 to improve torisional and VdW parameters using experimental 
heats of vaporization and densities, and validated
the overall result by predicting small-angle x-ray scattering 
results for model lipid bilayers to create the GROMOS 43-A1S3 lipid force field which gave excellent reproduction of SAXS form-factors for
different lipids~\cite{chiu:2009}.

Molecular dynamics can give this resolution while still reproducing the large-scale properties of the system as a whole~\cite{pandit:2008:simulationtextbook,dror:2012:biomicroscope}. 
Properties of the lipid bilayer can be expressed directly in terms of the positions or dynamics of atoms, 
providing additional understanding of these systems.

%There are many computational methods and models available to do simulations of systems of particles. In general, models
%must strike a balance between computational efficiency and accuracy of results; 
%due to computational limitations, large systems or long simulation runs require approximations be made in order 
%to be computationally feasible~\cite{pandit:2008:simulationtextbook}.
%
%Molecular dynamics evolves a system of particles by integrating the forces in a system of particles to calculate velocities according 
%to Newton's laws of motion~\cite{allen:2017:compsimliq}. 
%Physical properties that are needed for defining the 
%hamiltonian of the system are approximated and 
%defined in a ``force-field'', that is developed and optimized to reproduce experimental data~\cite{pandit:2008:simulationtextbook,chiu:2009,leonard:2019:developing}. 
%For our lipid force field, we utilize the gromos 43-A1S3 ``United Atom'' lipid force field, which coarse-grains
%hydrocarbons by removing all but the polar hydrogens and adding their mass and charge to the carbon species, which
%allows for a significant gain in speed over all-atom simulations~\cite{lyubartsev:2016:force}. Particles are
%expressed as points in space with a fixed mass and charge.
%Bonds and bond angles between particles in molecules are modelled as harmonic potentials, and dihedrals are modeled using periodic functions~\cite{pandit:2008:simulationtextbook}.
%Non-bonded interactions are treated using simple pairwise additive potentials for Van der Waals (VdW) and electrostatic interactions between particles.
%VdW interactions are approximated using the Lennard-Jones potential~\cite{gromacsmanual}, 
%\begin{equation}
%    V_{LJ}=\sum_{i<j}{\frac{C^{(12)}_{ij}}{r^{12}_{ij}}-\frac{C^{(6)}_{ij}}{r^{6}_{ij}}}
%\end{equation}
%between two particles~$i$~and~$j$. This potential gives particles a hard-shell radius as well as an attractive dispersion
%force. Electrostatic interactions are calculated using a coulomb potential~\cite{gromacsmanual}
%\begin{equation}
%    V_{ij}=f\frac{q_i q_j}{e_r r_{ij}}
%\end{equation}
%for two particles~$i$~and~$j$. This potential is used only for short range interactions --- long range electrostatics
%are computed using the particle-mesh Ewald method~\cite{gromacsmanual,pme:1995}.

\chapter{Interaction of salt with ether- and ester-linked phospholipid bilayers}
In this work we seek to answer the general questions of how salt inte\-racts 
with ether-linked lipid bilayers, and how ether-linked lipid bilayers differ from ester-linked lipid bilayers in the presence of salt.
We addressed these questions by carrying out a molecular dynamics simulation of 
1-hexadecyl-2-(9-octadecenyl)-sn-glycero-3-phosphatidylcholine lipid bilayer (HOPC) in NaCl salt. 
We also compared results of this simulation against our previous simulation of 
1-palmitoyl-2-oleoyl-sn-glycero-3-phosphatidylcholine lipid bilayer (POPC) 
in NaCl \cite{kruczek:2017}. Additionally, to gain insight into how salt modulates differences between ether- 
and ester-linked lipid bilayers, we also compared these simulations against those of ether- and ester-linked lipid bilayers in pure water~\cite{kruczek:2017:ether}. 
\section{Methods}
The trajectories for the POPC bilayer in 200 mM NaCl were taken from our previous work~\cite{kruczek:2017}, 
and were re-analyzed to address the specific goals of this project. The trajectories
for the HOPC bilayer were generated as described below, following the same protocol we used for the POPC bilayer. 
Analysis was carried out using GROMACS~\cite{abraham:2015,pall:2014,van:2005,lindahl:2001,berendsen:1995} utility 
tools, and in-house software developed using GROMACS API.

\subsection{HOPC bilayer construction}
To construct the HOPC bilayer, we first place 100 lipids on $10 \times 10$ nm grid to form a bilayer leaflet, and then reflect the leaflet to 
create the second leaflet of the bilayer. We then introduce 30,000 water molecules into the space outside the bilayer, which gives a
 150:1 waters to lipid ratio. We then replace 216 randomly-selected waters with 108 \na and 108 \cl ions to set an initial 
NaCl concentration of 200 mM. Note that the high water to lipid ratio is necessary to get a good representation of bulk solvent, 
as the addition of salt increases the distance of solvent ordering from the bilayer surface~\cite{kruczek:2017}. We then 
remove bad contacts by energy minimizing using the steepest descent approach. Finally, we subject the system to one cycle of annealing. Heating 
is done under constant pressure to 500K, but in steps of 100K to avoid excess kinetic energy divergence. During heating, each step is simulated 
for 10 ps. Cooling is also performed under constant pressure, but in smaller decrements -- cooling from 500K to 
400K is done in 5 steps, and cooling from 400K to 300K in steps of 10 steps. Each cooling step is simulated for 50 ps. The total annealing time is 850 ps.

\subsection{Simulation Details}
MD simulation is performed using GROMACS version 5.1.5~\cite{abraham:2015,pall:2014,van:2005,lindahl:2001,berendsen:1995}. Water is described using 
the SPC/E model~\cite{spce}, and we use the united atom Gromos 43A1-S3 parameters for lipids and 
lipid-water cross terms developed by our group~\cite{chiu:2009, kruczek:2017:ether}. 
These parameters have been verified in previous works without salt 
against experimental SAXS data
to ensure accurate reproduction of bilayer structure~\cite{kruczek:2017:ether}.
Parameters for HOPC are derived from those for ether-linkages used 
for di-hexadecyl phosphatidylcholine developed for previous work without ions~\cite{kruczek:2017:ether}.
Ion-ion and ion-water interactions are described using parameters developed by Joung and Cheatham III~\cite{joung:2008}, 
and the ion-lipid cross terms are calculated explicitly 
using Lorentz–Berthelot rules ~\cite{kruczek:2017}. These ion parameters
were verified in previous work for binding constants as well 
as electrostatic potential, and we reported good reproduction of experimental
results~\cite{kruczek:2017}. Temperature is held constant at 
300K using the Nos\'e-Hoover thermostat~\cite{nose:1983}, and a coupling constant of 0.5 ps. 
Pressure is maintained at 1 bar using the Parrinello-Rahman semisotropic barostat~\cite{parrinello:1981}, and 
a pressure coupling constant of 1.5 ps. 
All bonds are constrained using the P-LINCS algorithm~\cite{lincs}.
Neighbor-lists are updated every other time step, and integration is carried out using the leap-frog Verlet scheme. 
Long-range electrostatics beyond 16~\AA are computed using the smooth particle mesh Ewald algorithm~\cite{essmann:1995}, 
and Lennard-Jones interactions are calculated with a cutoff of 16~\AA.  
As before~\cite{kruczek:2017}, a continuous MD simulation is run for 0.5~$\mu$s.

\section{Results and Discussion}

\subsection{Bilayer Structure}

We first compare mass densities of the different chemical components of HOPC against those 
of POPC (see figure~\ref{fig:massdens}).

\begin{figure}[p]
    \caption[Comparison of component mass densities between the HOPC and POPC systems.]{ 
Comparison of component mass densities between the HOPC and POPC systems. 
Densities are computed by dividing the box dimension along the membrane normal into 2000 slices. For 
clairity of presentation, error bars are indicated for every 10th slice. 
Mass densities of solvent and ions, water, lipid headgroup and backbone, and lipid chains are shown. 
The plot for HOPC shows a peak in the water density inside the headgroup region that is not present in POPC, 
similar to what was seen in our previous work with DHPC~\cite{kruczek:2017:ether}. 
}
\label{fig:massdens}
\includegraphics[width=	0.7\textwidth]{dens_3component.eps}
\end{figure}
We find that the headgroup region of 
HOPC has a higher density of solvent compared to POPC -- this is visible as 
a distinct peak in the solvent density profile in the HOPC system. 
We verified that this additional peak does not
result from accumulation of ions by comparing mass density profiles of only water molecules. This
peak is similar to observations of a smaller peak 
made in our previous simulations in the absence of salt,
and may be a direct result of the ether-linked lipids~\cite{kruczek:2017:ether}.

Table~\ref{tab:struc} compares the structural properties of HOPC and POPC bilayers. 
Lipid volumes $V_l = V_c + V_{hg}$, where $V_c$ and $V_{hg}$ are 
\begin{table}
    \caption[Comparison of HOPC and POPC bilayer structure properties.]{ 
Comparison of HOPC and POPC bilayer structure properties: $V_l$ is the lipid
volume; $V_c$ and $V_h$ are the partial volumes of lipids chains and headgroups,
respectively; $2D_c$ and $D_b$ are chain and bilayer thickness, respectively; $A_l=V_c/D_c$ is the area per lipid.
$N_{w}$ is the number of perturbed waters in the system, defined as those behind the \emph{hydration boundary}. $\Delta \nu$ is the quadrupolar splitting constant of each system.}
\label{tab:struc}
\begin{tabularx}{\textwidth}{X|X|X}%{ c|c|c|c }
& POPC with NaCl & HOPC with NaCl \\ \hline
$V_c$ (\AA$^3$) & 898.4 $\pm$ 1.2 & 900.0 $\pm$ 1.0   \\
$V_h$ (\AA$^3$) & 318.9 $\pm$ 0.9 & 290.2 $\pm$ 0.9   \\
$V_l$ (\AA$^3$) & 1217.3 $\pm$ 0.5 & 1190.2 $\pm$ 0.9   \\
$2D_c$ (\AA) & 30.95 $\pm$ 0.32 & 32.13 $\pm$ 0.25   \\
$D_b$ (\AA) & 46.22 $\pm$ 0.49 & 46.05 $\pm$ 0.34   \\
$A_l$ (\AA$^2$) & 58.07 $\pm$ 0.63 & 56.03 $\pm$ 0.45   \\
$D_{hh}$ (\AA) & 40.24 $\pm$ 0.75 & 42.62 $\pm$ 0.88   \\
$N_{w}$ & 22.6 & 22.3  \\ 
$\Delta \nu$ (Hz) & 194.99 & 1756.08 \\
\end{tabularx}
\end{table}
the volumes of lipid chains (tails) and headgroups. 
These are computed using the approach given in 
Petrache et al.~\cite{petrache:1997}. In this approach, 
we first compute the number densities $n_i(z)$ of the various system 
components as a function of the bilayer normal ($z$). We then use these 
$n_i(z)$ to determine their partial molecular volumes, $v_i$, by optimizing the objective function, 
\begin{equation}
\label{eq:petrache}
\Omega(v_i)=\sum^{n_s}_{z_j}(1-\sum_{i=1}^{N_{\text{groups}}}(n_i(z_j)v_i))^2.
\end{equation}
Here, $N_{\text{groups}}=7$ are the different system 
components including methyls ($v_{CH_3}$), methines ($v_{CH_2}$), 
methylenes ($v_{CH}$), headgroup carbons, headgroup and backbone oxygens, headgroup phosphate
and nitrogen, and water ($v_{H_2O}$). 
%These partial volumes are shown in supplemental figure~\ref{supp:tab:volumes}.
Using these partial molecular volumes, we compute 
$V_c = 2v_{CH_3} + 2v_{CH_1} + 28v_{CH_2}$ and $V_{hg} = 10v_{hg_C} + 2v_{P\text{\&}N} + 8v_{hg_O}$. 
We find that the headgroup volume of HOPC is 
27.1~\AA$^3$~smaller than POPC, due likely to the absence of carbonyl groups. 
As expected for lipids with the same
hydrocarbon chains, the chain volumes are similar in the two lipids.

We also use the number densities computed above ($n_i(z)$) to determine lipid chain 
thickness ($2D_c$) and bilayer thickness ($D_b$). These results are found experimentally from 
x-ray and neutron scattering~\cite{fogarty:2015}. In our simulations these parameters 
are found by first computing the probability of
finding the $i$-th group in each slice of the simulation box,
$P_i(z)={n_i(z)}/{\sum_i n_i(z)}$. We define $2D_c$ as the distance between the Gibbs' surfaces on the 
interfacial probability densities of lipid chains. Each of the two surfaces is computed by identifying the
point on the curve where 
the integrals of probability densities in the interfacial regions above and below the surface are equal. 
We compute $D_b$ in a similarly, but from solvent probability densities. 
We find that both POPC and HOPC have similar membrane thicknesses ($D_b$); however a comparison of $2D_c$ 
indicates that the hydrophobic chains in HOPC are slightly thicker than those in POPC, reflecting tighter chain packing. This 
is also evident from comparison of electron densities (see figure~\ref{fig:eledens}).

\begin{figure}[p]
    \caption[Electron Densities of Simulated Systems.]{ 
Electron Densities of Simulated Systems. Symmetrized electron density distribution of each simulated system as a 
function of $z$. We can see the distribution for HOPC 
is broader, with a longer peak-to-peak distance than in the POPC distribution. Furthermore, 
we see a deeper trough in the middle of the HOPC distribution. This plot was made by using the gromacs densities tool, 
dividing the system into 2000 slices. This data was calculated for every five 
nanoseconds, and averaged over the last 150ns of trajectories. Errorbars are shown for every 10th data point.
}
\label{fig:eledens}
\includegraphics[width=	0.7\textwidth]{ele_scattering_density.eps}
\end{figure}
Electron density distributions are experimentally measured using x-ray scattering, by taking the reverse-fourier transform of
the scattering form-factor~\cite{fogarty:2015,nagle:2000,pandit:2008:simulationtextbook}. Since the experimental x-ray scattering is dependent on the model used to
compute the reverse-transform, we often calculate the form-factor of our simulated bilayer to compare to experiment by transforming the
electron density distribution~\cite{chiu:2009,kruczek:2017:ether,kruczek:2017}.
Since such data is not available for our current lipid system as far as we know, we cannot do this comparison here. 
%We have included this data in supplementary materials in case
%a comparsion becomes possible in the future (see supplemental figure~\ref{supp:formfactor}).

We also use the electron denisty distribution to calculate the peak-to-peak distance $D_{hh}$, 
which is used another measure of the bilayer thickness. This is roughly reflective of the distance between the 
electron-dense phosphate groups on either bilayer leaflet.

$D_{hh}$ values for POPC and HOPC indicate that the ether-linked bilayer is \about2\AA thicker on average than the ester-linked bilayer. 
This discrepancy between the values of $D_b$ and $D_{hh}$ is likely due to the irregular shape of the 
distribution of solvent in the headgroup of HOPC --- the extra region 
of solvent accumulation may move the Gibbs' surface used for finding $D_b$ into the bilayer surface.

The difference in bilayer thickness can also be shown directly from comparison of chain order 
parameters (See figure~\ref{fig:chainorder}). As lipid chains become more ordered on average, bilayer thickness
increases~\cite{nagle:2000}.
